---
title: "Post-Count QC"
author: "CCBR RNAseq pipeline"
output: html_document
params:
  folder: result
  sampleinfo: sampletable.txt
  data: RawCountFile_genes_filtered.txt
  projectId: "<projectId>"
  projectDesc: "<desc>"
---

```{r, include=FALSE, warning=FALSE, message=FALSE}
countMethod <- "subread"
if(grepl("rsem",params$data)){
  countMethod <- "rsem"
}
projectId<-params$projectId
projectDesc<-params$projectDesc
dateandtime<-format(Sys.time(), "%a %b %d %Y - %X")
```
### **Project:**
####    *`r projectId`*
### **Description:** 
####    *`r projectDesc`*
###  **Count method:** 
####    *`r countMethod`* 
### **Report generated:** 
####    *`r dateandtime`* 


```{r setup, echo=FALSE, warning=FALSE,message=FALSE,fig.align='center'}

library(rgl)
library(knitr)
library('edgeR')
library('statmod')
library('RColorBrewer') 
library('gplots')
library('reshape') 
library('ggplot2')
library('ggfortify')
library(amap)
library(DT)
library(plotly)
library('geneplotter')
library('DESeq2')

knit_hooks$set(rgl = function(before, options, envir) {
  if (!before) {
    ## after a chunk has been evaluated
    if (rgl.cur() == 0) return()  # no active device
    name = paste(options$fig.path, options$label, sep = '')
    rgl.snapshot(paste(name, '.png', sep = ''), fmt = 'png')
    return(paste('\\includegraphics{', name, '}\n', sep = ''))
  }
})

knit_hooks$set(webgl = hook_webgl)
```

```{r, echo=FALSE, warning=FALSE,message=FALSE}

## grab args & normalization

DIR <- params$folder
FILE1 <- params$sampleinfo
FILE2 <- params$data

# ANNOTATE <- args[4]

#
## setwd(DIR)
# read files
sampleinfo=read.delim(FILE1)
x = read.delim(FILE2,row.names=1)
colnames(x)=as.character(sampleinfo[,4])
sampleFiles=as.character(sampleinfo[,2])
## read annotation file
## ann=read.delim(ANNOTATE)
# DGElist object --------------------------------------------------------------
condition = as.factor(sampleinfo$condition)
y = DGEList(counts=x,group=condition)
## Normalization TMM ------------------------------------------------------------
## method = =c("TMM","RLE","upperquartile","none")
y <- calcNormFactors(y,method="TMM")
# y$samples

#png("libdistrib.png")
#par(mfrow=c(1,4))
#barplot(y$samples$lib.size*1e-6,main="Before normalization", names= strsplit(colnames(y$counts),".star.count.txt"), ylab="Library size (millions)",las=2,cex.names=0.8)
#barplot(y$samples$lib.size*y$samples$norm.factors*1e-6,main="TMM", names= strsplit(colnames(y$counts),".star.count.txt"), ylab="Library size (millions)",las=2,cex.names=0.8)
ddsHTSeq<-DESeqDataSetFromMatrix(countData=x,colData=sampleinfo, design=~condition)
dds<-DESeq(ddsHTSeq)
dds.ndata=as.data.frame(counts(dds,normalized=TRUE))
#barplot(colSums(dds.ndata)*1e-6,main="DESeq2", names= strsplit(colnames(y$counts),".star.count.txt"), ylab="Library size (millions)",las=2,cex.names=0.8)
Group <- factor(sampleinfo$condition)
design=model.matrix(~0+Group)
v1 <- voom(as.matrix(x),design,plot=FALSE,normalize="quantile")
#barplot(colSums(2^(v1$E))*1e-6,main="Limma", names= strsplit(colnames(y$counts),".star.count.txt"), ylab="Library size (millions)",las=2,cex.names=0.8)
#dev.off()


## estimating common and tagwise dispersions -----------------------------------------
y <- estimateCommonDisp(y)
y <- estimateTagwiseDisp(y) #default trend: moveingave

ylog2=cpm(y,log=TRUE,normalized.lib.sizes=TRUE,prior.count=0.5) # prior count like avelogcpm
rawlog2= cpm(y,log=TRUE,normalized.lib.sizes=FALSE,prior.count=0.5)
ddslog2= cpm(dds.ndata,log=TRUE,normalized.lib.sizes=FALSE,prior.count=0.5) 
## save it
```

## **Count Distribution** {.tabset}

### Before Normalization

```{r, echo=FALSE, warning=FALSE,message=FALSE}
print(ggplot(melt(as.data.frame(rawlog2))) + geom_density(aes(x = value,colour = variable)) + labs(x = NULL) + theme(legend.position='right') + scale_x_log10())
```

### TMM

```{r, echo=FALSE, warning=FALSE,message=FALSE}
print(ggplot(melt(as.data.frame(ylog2))) + geom_density(aes(x = value,colour = variable)) + labs(x = NULL) + theme(legend.position='right') + scale_x_log10())
```

### DESeq2

```{r, echo=FALSE, warning=FALSE,message=FALSE}
print(ggplot(melt(as.data.frame(ddslog2))) + geom_density(aes(x = value,colour = variable)) + labs(x = NULL) + theme(legend.position='right') + scale_x_log10())
```

### Limma

```{r, echo=FALSE, warning=FALSE,message=FALSE}
print(ggplot(melt(as.data.frame(v1$E))) + geom_density(aes(x = value,colour = variable)) + labs(x = NULL) + theme(legend.position='right') + scale_x_log10())
```

## **PCA Plots** {.tabset}

### Before Normalization

```{r, echo=FALSE,webgl=TRUE,message=FALSE, warning=FALSE}

## PCA for before norm

before.edf=rawlog2
before.tedf= t(before.edf)
before.pca=prcomp(before.tedf,scale.=T)
before.tedf1 = data.frame(before.tedf)
Phenotype=sampleinfo$condition
cell_rep=sampleinfo$label
before.tedf1$group = as.factor(Phenotype)
before.pc1 = round(before.pca$sdev[1]^2/sum(before.pca$sdev^2)*100,2)
before.pc2 = round(before.pca$sdev[2]^2/sum(before.pca$sdev^2)*100,2)
before.pc3 = round(before.pca$sdev[3]^2/sum(before.pca$sdev^2)*100,2)


# plot(before.pca,type="lines")  #Decide how many PC's are relevant for plotting
  #before.pca$x[,1:3]  #look at first 3 PC's

plot3d(before.pca$x[,1:3],col = as.integer(before.tedf1$group),type="s",size=2,main="PCA before normalization",xlab=paste0("PC1 (",before.pc1,"%)"),ylab=paste0("PC2 (",before.pc2,"%)"),zlab=paste0("PC3 (",before.pc3,"%)"))
group.v<-as.vector(cell_rep)
text3d(before.pca$x, before.pca$y, before.pca$z, group.v, cex=1.0, adj = 1.2)
legend3d("topright", legend = levels(sampleinfo$condition), pch = 16, col = as.numeric(as.factor(levels(sampleinfo$condition))), cex=0.5)
#rgl.postscript("pca3d_raw.pdf","pdf")
rgl.snapshot("pca3d_raw.png","png")

```

### TMM

```{r, echo=FALSE,webgl=TRUE,message=FALSE, warning=FALSE}

## PCA for edgeR

edgeR.edf=ylog2
edgeR.tedf= t(edgeR.edf)
edgeR.pca=prcomp(edgeR.tedf,scale.=T)
edgeR.tedf1 = data.frame(edgeR.tedf)
Phenotype=sampleinfo$condition
cell_rep=sampleinfo$label
edgeR.tedf1$group = as.factor(Phenotype)
edgeR.pc1 = round(edgeR.pca$sdev[1]^2/sum(edgeR.pca$sdev^2)*100,2)
edgeR.pc2 = round(edgeR.pca$sdev[2]^2/sum(edgeR.pca$sdev^2)*100,2)
edgeR.pc3 = round(edgeR.pca$sdev[3]^2/sum(edgeR.pca$sdev^2)*100,2)


# plot(edgeR.pca,type="lines")  #Decide how many PC's are relevant for plotting
  #edgeR.pca$x[,1:3]  #look at first 3 PC's

plot3d(edgeR.pca$x[,1:3],col = as.integer(edgeR.tedf1$group),type="s",size=2,main="PCA after TMM normalization",xlab=paste0("PC1 (",edgeR.pc1,"%)"),ylab=paste0("PC2 (",edgeR.pc2,"%)"),zlab=paste0("PC3 (",edgeR.pc3,"%)"))
group.v<-as.vector(cell_rep)
text3d(edgeR.pca$x, edgeR.pca$y, edgeR.pca$z, group.v, cex=1.0, adj = 1.2)
legend3d("topright", legend = levels(sampleinfo$condition), pch = 16, col = as.numeric(as.factor(levels(sampleinfo$condition))), cex=0.5)
#rgl.postscript("pca3d_edgeR.pdf","pdf")
rgl.snapshot("pca3d_edgeR.png","png")

```

### DESeq2

```{r, echo=FALSE,webgl=TRUE,message=FALSE, warning=FALSE}

## PCA for deseq2

#rld <- rlogTransformation(dds, blind=TRUE)
#rldm=assay(rld)
#colnames(rldm)=colnames(x)
#deseq2.edf=as.matrix(rldm)
deseq2.edf=ddslog2
deseq2.tedf= t(deseq2.edf)
deseq2.tedf=deseq2.tedf[,apply(deseq2.tedf,2,var)!= 0]
deseq2.pca=prcomp(deseq2.tedf,scale.=T)
deseq2.tedf1 = data.frame(deseq2.tedf)
Phenotype=sampleinfo$condition
cell_rep=sampleinfo$label
deseq2.tedf1$group = as.factor(Phenotype)
deseq2.pc1 = round(deseq2.pca$sdev[1]^2/sum(deseq2.pca$sdev^2)*100,2)
deseq2.pc2 = round(deseq2.pca$sdev[2]^2/sum(deseq2.pca$sdev^2)*100,2)
deseq2.pc3 = round(deseq2.pca$sdev[3]^2/sum(deseq2.pca$sdev^2)*100,2)


plot3d(deseq2.pca$x[,1:3],col = as.integer(deseq2.tedf1$group),type="s",size=2,main="PCA after DESeq2 normalization",xlab=paste0("PC1 (",deseq2.pc1,"%)"),ylab=paste0("PC2 (",deseq2.pc2,"%)"),zlab=paste0("PC3 (",deseq2.pc3,"%)"))
group.v<-as.vector(cell_rep)
text3d(deseq2.pca$x, deseq2.pca$y, deseq2.pca$z, group.v, cex=1.0, adj = 1.2)
legend3d("topright", legend = levels(sampleinfo$condition), pch = 16, col = as.numeric(as.factor(levels(sampleinfo$condition))), cex=0.5)
#rgl.postscript("pca3d_deseq2.pdf","pdf")
rgl.snapshot("pca3d_deseq2.png","png")

```

### Limma

```{r, echo=FALSE,webgl=TRUE,message=FALSE, warning=FALSE}

## PCA for Limma

limma.edf=as.matrix(v1$E)
limma.tedf= t(limma.edf)
limma.tedf=limma.tedf[,apply(limma.tedf,2,var)!= 0]
limma.pca=prcomp(limma.tedf,scale.=T)
limma.tedf1 = data.frame(limma.tedf)
Phenotype=sampleinfo$condition
cell_rep=sampleinfo$label
limma.tedf1$group = as.factor(Phenotype)
limma.pc1 = round(limma.pca$sdev[1]^2/sum(limma.pca$sdev^2)*100,2)
limma.pc2 = round(limma.pca$sdev[2]^2/sum(limma.pca$sdev^2)*100,2)
limma.pc3 = round(limma.pca$sdev[3]^2/sum(limma.pca$sdev^2)*100,2)


plot3d(limma.pca$x[,1:3],col = as.integer(limma.tedf1$group),type="s",size=2,main="PCA after Limma normalization",xlab=paste0("PC1 (",limma.pc1,"%)"),ylab=paste0("PC2 (",limma.pc2,"%)"),zlab=paste0("PC3 (",limma.pc3,"%)"))
group.v<-as.vector(cell_rep)
text3d(limma.pca$x, limma.pca$y, limma.pca$z, group.v, cex=1.0, adj = 1.2)
legend3d("topright", legend = levels(sampleinfo$condition), pch = 16, col = as.numeric(as.factor(levels(sampleinfo$condition))), cex=0.5)
#rgl.postscript("pca3d_limma.pdf","pdf")
rgl.snapshot("pca3d_limma.png","png")

```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
before.dfm <- melt(as.data.frame(rawlog2))
edgeR.dfm <- melt(as.data.frame(ylog2))
deseq2.dfm <- melt(as.data.frame(ddslog2))
limma.dfm <- melt(as.data.frame(v1$E))
```

## **Boxplots** {.tabset}

### Before Normalization

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE}
par(mar=c(par("mar")[1]+5,par("mar")[-1]))
boxplot(value~variable,las=2,data=before.dfm,main="Before normalization",ylab="Counts",col=as.numeric(as.factor(sampleinfo$condition)))
```

### TMM

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE}
par(mar=c(par("mar")[1]+5,par("mar")[-1]))
boxplot(value~variable,las=2,data=edgeR.dfm,main="TMM",ylab="Counts",col=as.numeric(as.factor(sampleinfo$condition)))
```

### DESeq2

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE}
par(mar=c(par("mar")[1]+5,par("mar")[-1]))
boxplot(value~variable,las=2,data=deseq2.dfm,main="DESeq2",ylab="Counts",col=as.numeric(as.factor(sampleinfo$condition)))
```

### Limma

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE}
par(mar=c(par("mar")[1]+5,par("mar")[-1]))
boxplot(value~variable,las=2,data=limma.dfm,main="Limma",ylab="Counts",col=as.numeric(as.factor(sampleinfo$condition)))
```

## **Similarity Heatmaps Using Pearson - Complete Linkage** {.tabset}

```{r, echo=FALSE,message=FALSE, warning=FALSE}
hmapheight = length(sampleinfo$label)
if(hmapheight<8){
hmapheight = 8
}
```

### Before Normalization

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fid.width=500, fig.height=hmapheight}

hmcol <- colorRampPalette(c("black","red","yellow","white"),space="rgb")(100)
before.distrawlog2=amap::Dist(t(rawlog2),method="pearson")
before.mat = as.matrix(before.distrawlog2)
heatmap.2(before.mat, trace="none", col = rev(hmcol), labCol=FALSE, colRow=as.numeric(as.factor(sampleinfo$condition)), margin=c(16, 16), main="Before normalization")
```

### TMM

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fid.width=500, fig.height=hmapheight}

edgeR.distylog2=amap::Dist(t(ylog2),method="pearson")
edgeR.mat = as.matrix(edgeR.distylog2)
heatmap.2(edgeR.mat, trace="none", col = rev(hmcol), labCol=FALSE, colRow=as.numeric(as.factor(sampleinfo$condition)), margin=c(16, 16), main="TMM")
```

### DESeq2

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fid.width=500, fig.height=hmapheight}

deseq2.dists <- amap::Dist(t(ddslog2),method="pearson")
deseq2.mat <- as.matrix(deseq2.dists)
heatmap.2(deseq2.mat, trace="none", col = rev(hmcol), labCol=FALSE, colRow=as.numeric(as.factor(sampleinfo$condition)), margin=c(16, 16), main="DESeq2")
```

### Limma

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fid.width=500, fig.height=hmapheight}

limma.dists <- amap::Dist(t(limma.edf),method="pearson")
limma.mat <- as.matrix(limma.dists)
heatmap.2(limma.mat, trace="none", col = rev(hmcol), labCol=FALSE, colRow=as.numeric(as.factor(sampleinfo$condition)), margin=c(16, 16), main="Limma")
```


## **MD Plots** (sample by sample mean vs fold change of expression)

```{r, echo=FALSE,message=FALSE,fig.show='hold',fig.align='center', warning=FALSE, fid.width=200, fig.height=3.3*length(sampleinfo$label)}
par(mfcol=c(length(sampleinfo$label),4))
for(i in 1:length(sampleinfo$label)){
  plotMD(rawlog2,column=i, main=paste0("Raw ",sampleinfo$label[i]), xlim=c(-5,15), ylim=c(-15,15))
  abline(h=0, col="red", lty=2, lwd=2)
}
for(i in 1:length(sampleinfo$label)){
  plotMD(ylog2,column=i, main=paste0("TMM ",sampleinfo$label[i]), xlim=c(-5,15), ylim=c(-15,15))
  abline(h=0, col="red", lty=2, lwd=2)
}
for(i in 1:length(sampleinfo$label)){
  plotMD(ddslog2,column=i, main=paste0("DESeq2 ",sampleinfo$label[i]), xlim=c(-5,15), ylim=c(-15,15))
  abline(h=0, col="red", lty=2, lwd=2)
}
for(i in 1:length(sampleinfo$label)){
  plotMD(v1$E,column=i, main=paste0("Limma ",sampleinfo$label[i]), xlim=c(-5,15), ylim=c(-15,15))
  abline(h=0, col="red", lty=2, lwd=2)
}
```