---
title: "Inital/QC Report"
author: "CCBR scRNAseq pipeline"
output: html_document
params:
  matrix: "path/to/count/matrix"
  species: "hg19"
  projectId: "<projectId>"
  projectDesc: "<desc>"
---

```{r headers, include=FALSE, warning=FALSE, message=FALSE}
projectId<-params$projectId
projectDesc<-params$projectDesc
dateandtime<-format(Sys.time(), "%a %b %d %Y - %X")
species<-params$species
matrix<-params$matrix
```

### **Project:**
####    *`r projectId`*
### **Description:** 
####    *`r projectDesc`*
### **Report generated:** 
####    *`r dateandtime`* 

```{r setup, echo=FALSE, warning=FALSE,message=FALSE,results='hide',fig.keep='all'}
# library(rgl)
library(knitr)
library(Seurat)
library(dplyr)
library(Matrix)
library(scater)
library(scran)
library(limma)
library(GSVA)
library(GSEABase)
library(genefilter)
library(org.Hs.eg.db)
library(gplots)
library(gdata)

# knit_hooks$set(rgl = function(before, options, envir) {
#   if (!before) {
#     ## after a chunk has been evaluated
#     if (rgl.cur() == 0) return()  # no active device
#     name = paste(options$fig.path, options$label, sep = '')
#     rgl.snapshot(paste(name, '.png', sep = ''), fmt = 'png')
#     return(paste('\\includegraphics{', name, '}\n', sep = ''))
#   }
# })

#knit_hooks$set(webgl = hook_webgl)

sce <- read10XResults(data_dir = matrix, min_total_cell_counts = 0, min_mean_gene_counts = 0, expand = TRUE)

is.mito <- grepl("^MT-",fData(sce)$symbol)|grepl("^mt-",fData(sce)$symbol)
sce <- calculateQCMetrics(sce, feature_controls=list(Mt=is.mito))
libsize.drop <- isOutlier(sce$total_counts, nmads=3, type="lower", log=TRUE)
feature.drop <- isOutlier(sce$total_features, nmads=3, type="lower", log=TRUE)
mito.drop <- isOutlier(sce$pct_counts_feature_controls_Mt, nmads=3, type="higher")
sce$keep <- !(libsize.drop | feature.drop | mito.drop)
sce_filtered <- sce[,!(libsize.drop | feature.drop | mito.drop)]
numcells <- nexprs(sce_filtered, byrow=TRUE)
keepgenes <- numcells >= round(0.001*ncol(sce))
sce_filtered <- sce_filtered[keepgenes,]
numdroppedgenes <- sum(!keepgenes)
totalgenes <- length(keepgenes)
keepcells <- rownames(sce_filtered@phenoData)
droppedcells <- setdiff(rownames(sce@phenoData),keepcells)

so.data <- Read10X(matrix)
so <- new("seurat", raw.data=so.data)
so <- Setup(so, min.genes=0, min.cells=0, do.logNormalize=F, do.scale=F, do.center=F, project=projectId)
so <- SubsetData(so, cells.use=keepcells)
so@raw.data <- so@data
so <- Setup(so, min.genes=0, min.cells=round(0.001*ncol(sce)), do.logNormalize=T, do.scale=F, do.center=F, total.expr=1e4, project=projectId, save.raw=F)
```

### **Pre-Filter Histograms**

```{r histograms_1, echo=FALSE,warning=FALSE,message=FALSE}
par(mfrow=c(1,2))
hist(sce$total_counts, xlab="Library size", main="", 
    breaks=20, col="grey80", ylab="Number of cells")
plot(density(sce$total_counts),col="red", xlab="Library size", main="")
```
```{r histograms_2, echo=FALSE,warning=FALSE,message=FALSE}
par(mfrow=c(1,2))
hist(sce$total_features, xlab="Number of expressed genes", main="", 
    breaks=20, col="grey80", ylab="Number of cells")
plot(density(sce$total_features),col="red", xlab="Number of expressed genes", main="")
```
```{r histograms_3, echo=FALSE,warning=FALSE,message=FALSE}
par(mfrow=c(1,2))
hist(sce$pct_counts_feature_controls_Mt, xlab="Mitochondrial proportion (%)", 
    ylab="Number of cells", breaks=20, main="", col="grey80")
plot(density(sce$pct_counts_feature_controls_Mt),col="red", xlab="Mitochondrial proportion (%)", main="")
```

### **Removed** `r numdroppedgenes` **lowly represented genes from count matrix from** `r totalgenes` **inital total genes.**  

### **Removed the following** `r length(droppedcells)` **barcoded cells as poor quality:**  


```{r filter_metrics, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
kable(data.frame(ByLibSize=sum(libsize.drop), ByFeature=sum(feature.drop), ByMito=sum(mito.drop), Remaining=ncol(sce)))
cat("  \n")
cat(unlist(droppedcells))
cat("  \n")
```

### **QC PCA to Visualize Dropped Cells**

```{r qc_pca, echo=FALSE,warning=FALSE,message=FALSE}
sce$keep_by_defaults <- (
    # remove cells with unusual numbers of genes
    !sce$filter_on_total_features &
    # remove cells with unusual numbers molecules counted
    !sce$filter_on_total_counts &
    # < 80% mitochondrial
    !sce$filter_on_pct_counts_feature_controls_Mt &
    # controls shouldn't be used in downstream analysis
    !sce$is_cell_control
)
scater::plotPCA(sce,
                size_by = "total_features", 
                colour_by = "keep",
                shape_by = "keep_by_defaults",
                pca_data_input = "pdata",
                detect_outliers = FALSE,
                return_SCESet = FALSE)
sce<-sce_filtered
```

```{r rle_plot, echo=FALSE,message=FALSE,warning=FALSE,results='hide',fig.keep='all'}
### **Relative Log Expression (Before and After CPM Normalization)

#set_exprs(sce, "log2_counts") <- log2(counts(sce)+sce@logExprsOffset)
#qclust <- quickCluster(sce, min.size = 30)
#sce <- computeSumFactors(sce, sizes = 15, clusters = qclust)
#sce <- normalize(sce)
#plotRLE(sce,exprs_mats=list(Raw="log2_counts",CPM="exprs"),exprs_logged=c(TRUE,TRUE))
```

### **nGenes vs nUMI** (Useful for outlier detection)

```{r nUMI_plot, echo=FALSE, message=FALSE, warning=FALSE}
#xthresh = median(so@data.info$nUMI)+3*stats::mad(so@data.info$nUMI)
#ythresh = median(so@data.info$nGene)+3*stats::mad(so@data.info$nGene)
GenePlot(so, "nUMI", "nGene")
```

### **Cell Cycle Stage**

```{r cell_cycle, echo=FALSE,warning=FALSE,message=FALSE}
if(grepl("hg",species)|grepl("mm",species)){
  if(grepl("hg",species)){
    pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", package="scran"))
  }
  if(grepl("mm",species)){
    pairs <- readRDS(system.file("exdata", "mouse_cycle_markers.rds", package="scran"))
  }
  assignments <- cyclone(sce_filtered, pairs, gene.names=rownames(sce_filtered))
  plot(assignments$score$G1, assignments$score$G2M, xlab="G1 score", ylab="G2/M score", pch=16, 
    col = ifelse((assignments$score$G1>=0.5)&(assignments$score$G2M<0.5),'green','orange'))
  abline(h=0.5,col="red")
  abline(v=0.5,col="red")
  legend("topleft", "[G2/M]", bty="n", text.col="red")
  legend("topright", "[?]", bty="n", text.col="red")
  legend("bottomleft", "[S]", bty="n", text.col="red")
  legend("bottomright", "[G1]", bty="n", text.col="red")
}
sce$G1score <- assignments$score$G1
sce$G2Mscore <- assignments$score$G2M
```

### **PCA 1:2 Before Regressing Cell Cycle Stage**

```{r regress_confounding, echo=FALSE,warning=FALSE,message=FALSE,message=FALSE,results='hide',fig.keep='all'}
norm_exprs(sce) <- removeBatchEffect(exprs(sce), covariates=data.frame(assignments$score[,c("G1", "G2M")],percent.mito=sce$pct_counts_feature_controls_Mt,nUMI=so@data.info$nUMI))
stage <- c()
for(i in 1:length(sce$G1score)){
  if(sce$G1score[i] >= 0.5 & sce$G2Mscore[i] >= 0.5){
    stage[i] <- "???"
  }else if(sce$G1score[i] >= 0.5){
    stage[i] <- "G1"
  }else if(sce$G2Mscore[i] >= 0.5){
    stage[i] <- "G2M"
  }else{
    stage[i] <- "S"
  }
}
metadata <- data.frame(percent.mito=sce$pct_counts_feature_controls_Mt,G1.score=sce$G1score,G2M.score=sce$G2Mscore,stage=stage,row.names=colnames(sce))
so <- AddMetaData(so, metadata)
so_temp <- RegressOut(so, latent.vars = c("nUMI", "percent.mito"))
so_temp <- MeanVarPlot(so_temp ,fxn.x = expMean, fxn.y = logVarDivMean, x.low.cutoff = 0.0125, x.high.cutoff = 3, y.cutoff = 0.5, do.contour = F, do.plot=F)
so_temp <- PCA(so_temp, pc.genes = so_temp@var.genes, do.print = FALSE)
so_temp <- ProjectPCA(so_temp,do.print=F)
PCAPlot(so_temp, 1, 2, group.by="stage",cols.use=c(rgb(1, 0, 0, 0.7),rgb(0, 1, 0, 0.3),rgb(0, 0, 1, 0.7),rgb(0, 0, 0, 0.7)))
so <- RegressOut(so, latent.vars = c("nUMI", "percent.mito", "G1.score", "G2M.score"))
```

### **PCA 1:2 After Regressing by G1/G2M Scores**

```{r regress_confounding_final, echo=FALSE,warning=FALSE,message=FALSE,message=FALSE,results='hide',fig.keep='all'}
so <- RegressOut(so, latent.vars = c("nUMI", "percent.mito", "G1.score", "G2M.score"))
so <- MeanVarPlot(so ,fxn.x = expMean, fxn.y = logVarDivMean, x.low.cutoff = 0.0125, x.high.cutoff = 3, y.cutoff = 0.5, do.contour = F, do.plot=F)
so <- PCA(so, pc.genes = so@var.genes, do.print = FALSE)
so <- ProjectPCA(so,do.print=F)
PCAPlot(so, 1, 2, group.by="stage",cols.use=c(rgb(1, 0, 0, 0.7),rgb(0, 1, 0, 0.3),rgb(0, 0, 1, 0.7),rgb(0, 0, 0, 0.7)))
```

### **Highest Expression**

```{r highestexpression, echo=FALSE,warning=FALSE,message=FALSE}
fontsize <- theme(axis.text=element_text(size=12), axis.title=element_text(size=16))
make.names(featureData(sce)$symbol,unique=TRUE)->featureNames(sce)
plotQC(sce, type = "highest-expression", n=50) + fontsize
featureData(sce)$id->featureNames(sce)
```

### **scater PC Metrics**

```{r findpcs, echo=FALSE,warning=FALSE,message=FALSE}
plotQC(sce, type = "find-pcs") + fontsize
```

### **Seurat MeanVar Plot**

```{r Seurat_meanvar, echo=FALSE,warning=FALSE,message=FALSE,results='hide',fig.keep='all'}
MeanVarPlot(so ,fxn.x = expMean, fxn.y = logVarDivMean, x.low.cutoff = 0.0125, x.high.cutoff = 3, y.cutoff = 0.5, do.contour = F)
```

### **Seurat PCA Genes**

```{r Seurat_vizpca, echo=FALSE,warning=FALSE,message=FALSE,results='hide',fig.keep='all',fig.height=70}
VizPCA(so, 1:20, nCol=2)
```

### **Seurat PC Heatmaps**

```{r Seurat_heatmaps_1, echo=FALSE,warning=FALSE,message=FALSE}
PCHeatmap(so, pc.use = 1:2, cells.use = 500, do.balanced = TRUE, label.columns = FALSE, use.full = FALSE)
```

```{r Seurat_heatmaps_2, echo=FALSE,warning=FALSE,message=FALSE}
PCHeatmap(so, pc.use = 3:5, cells.use = 500, do.balanced = TRUE, label.columns = FALSE, use.full = FALSE)
```

```{r Seurat_heatmaps_3, echo=FALSE,warning=FALSE,message=FALSE}
PCHeatmap(so, pc.use = 6:8, cells.use = 500, do.balanced = TRUE, label.columns = FALSE, use.full = FALSE)
```

```{r Seurat_heatmaps_4, echo=FALSE,warning=FALSE,message=FALSE}
PCHeatmap(so, pc.use = 9:11, cells.use = 500, do.balanced = TRUE, label.columns = FALSE, use.full = FALSE)
```

```{r Seurat_heatmaps_5, echo=FALSE,warning=FALSE,message=FALSE}
PCHeatmap(so, pc.use = 12:14, cells.use = 500, do.balanced = TRUE, label.columns = FALSE, use.full = FALSE)
```

```{r Seurat_heatmaps_6, echo=FALSE,warning=FALSE,message=FALSE}
PCHeatmap(so, pc.use = 15:17, cells.use = 500, do.balanced = TRUE, label.columns = FALSE, use.full = FALSE)
```

```{r Seurat_heatmaps_7, echo=FALSE,warning=FALSE,message=FALSE}
PCHeatmap(so, pc.use = 18:20, cells.use = 500, do.balanced = TRUE, label.columns = FALSE, use.full = FALSE)
```

### **Seurat PCElbowPlot**

```{r Seurat_elbow, echo=FALSE,warning=FALSE,message=FALSE}
PCElbowPlot(so)
```

```{r Seurat_save_rds, echo=FALSE,warning=FALSE,message=FALSE}
gdata::keep(so, params, species, projectId, sure=TRUE) ## memory optimizations
saveRDS(so, paste0(projectId,"_seurat_object.rds")) ## output SO
exprs <- as.matrix(so@data) ## only need expression matrix for next step
gdata::keep(exprs, params, species, sure=TRUE)
```

### **GSVA Heatmap**

```{r gsva_heatmap, echo=FALSE,warning=FALSE,message=FALSE,results='hide',fig.keep='all',fig.height=10}
if(grepl("hg",species)|grepl("mm",species)){
  eset <- ExpressionSet(assayData=exprs)
  eset_filtered <- varFilter(eset,var.func=IQR, var.cutoff=0.5, filterByQuantile=TRUE)
  if(grepl("hg",species)){
    gsets <- getGmt("/data/ngs/public/msigdb/msigdb.v4.0.symbols.gmt")
  }
  if(grepl("mm",species)){
    gsets <- getGmt("/data/ngs/public/msigdb/msigdb.v4.0.symbols.gmt")
  }
  scores <- gsva(exprs(eset_filtered),gsets,,min.sz=10,max.sz=500,verbose=TRUE)
  myscores <- scores$es.obs[unlist(lapply(rownames(scores$es.obs),function(x){grepl("^REACTOME_",x)|grepl("^KEGG_",x)|grepl("^BIOCARTA_",x)})),]
  top <- myscores[names(head(rev(sort((rowSds(myscores)))),50)),]
  heatmap.2(top,trace="none",labRow=NULL,margins=c(1,20),cexRow=0.7)
}
```
